---
title: "01_Create_Fragment_Profiles_update"
author: "Judy Ann Cocadiz"
date: "2023-04-02"
output:
  pdf_document: default
  html_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```
## Summary

* __Introduction:__*
+ Quick paragraph on logic behind this script

* __Setup:__*
+ Loads packages
+ Prepares sample info


* __Generate Fragment Profiles: __ *
+ Function to produce fragment profiles
+ Get vector of the sample names
+ Loop to produce fragment profile for each sample
+ Function to produce combined graph of fragment profiles
+ Creates combined graph for each sample


## Introduction

This is an update to creating fragment profiles so it is more generalisable.
To run this script you will need:
  - a csv file containing:
    * sample_id
    * bam_path (i.e. path/to/bam)
    * group (eg. high_gc, low_gc, healthy, delfi_gc)
  - output folder path


## Setup

```{r Setup}
library(here)
library(GenomicAlignments)
library(BSgenome.Hsapiens.UCSC.hg19)
library(ggplot2)
library(tidyverse)
library(GenomicRanges)
```


```{r enter sample info}
#sample_info <- "path/to/sample_data.tsv"
sample_info <- "~/Documents/GCproject_final/data/bams/all_sample_info.csv"
#output <- "path/to/output"
output <-  "~/Documents/GCproject_final/output/fragment_profiles"
#path to regions of genome to filter out eg. centromeres, low map regions
filter_path <- "/Volumes/archive/userdata/student_users/judyanncocadiz/Documents/GCproject_final/data/fragment_profiles/filters.hg19.rda"
```



```{r load sample info}
sample_info <- read_csv(sample_info)
samples <- sample_info$sample_id
bams <- sample_info$bam_path
samples
```



```{bash create folders to organise output, eval = FALSE}
# change to whatever your output directory is
# cd $output
mkdir granges
mkdir counts
mkdir graphs
```


It takes a long time to read in bams as GRange, so reading in and saving as RDS file for later use.
Set eval = TRUE if this has not been done for your samples
```{r Load bams as GRange function, eval = FALSE}

load_bam_as_GRange <- function(sample, bam_path, save = TRUE, save_path){
  print(paste("Loading", sample, "as GRange."))
  param <- ScanBamParam(flag = scanBamFlag(isDuplicate = FALSE,
                                           isSecondaryAlignment = FALSE,
                                           isUnmappedQuery = FALSE),
                        mapqFilter = 30)
  
  gp <- readGAlignmentPairs(bam_path, param = param)
  gr <- granges(keepSeqlevels(gp, paste0("chr", 1:22), pruning.mode="coarse"),
                on.discordant.seqnames="drop")
  if(save == TRUE){
    print("Saving GRange")
    saveRDS(gr, save_path)
  }
}

```


```{r load bams, eval = FALSE}
for(i in 1:length(samples)){
  save_path <- paste0(output, "/granges/", samples[i], ".gr.RDS")
  load_bam_as_GRange(samples[i], bam_path = bams[i], save_path = save_path)
}

```



Counts number of short and long frags in 5Mbp windows getting fragment profiles data
This can also take a while, so again only set eval = TRUE if this has not been done for your samples.
```{r function get fragment counts function, eval = FALSE}

fragment_counts <- function(sample_id, gr, output, filters, plot_density = TRUE){
  w.all <- width(gr)
  q.all <- quantile(w.all, c(0.001, 0.999))
  gr <- gr[which(w.all > q.all[1] & w.all < q.all[2])]
  w.all <- width(gr)
  gr.clean <- gr[-queryHits(findOverlaps(gr, filters.hg19))]
  lengths <- data.frame(length = w.all)
  
  if(plot_density == TRUE){
    uw <- unique(w.all)
    modal_width <- uw[which.max(tabulate(match(w.all, uw)))]
    dp <- ggplot(lengths, aes(x = length)) +
      geom_density() +
      geom_vline(xintercept = modal_width, linewidth = 0.2)
    ggsave(filename = paste0(output, "/graphs/", sample_id, ".dp.png"), plot = dp, device = "png")
  }
  
  print("Filtering")
  w.all <- width(gr.clean)
  fragments <- gr.clean[which(w.all >= 100 & w.all <= 220)]
  w <- width(fragments)
  bins <- tileGenome(seqinfo(Hsapiens),
                     tilewidth = 5000000,
                     cut.last.tile.in.chrom = TRUE)
  chrms <- paste0("chr", 1:22)
  bins <- bins[seqnames(bins) %in% chrms]
  
  print("Counting Fragments in each bin")
  frag.list <- split(fragments, w)
  #count fragments of each length for each bin
  counts <- sapply(frag.list, function(x) countOverlaps(bins, x)) 
  #put counts into matrix
  if(min(w) > 100) {
    m0 <- matrix(0, ncol=min(w) - 100, nrow=nrow(counts),
                 dimnames=list(rownames(counts), 100:(min(w)-1)))
    counts <- cbind(m0, counts)
  }
  
  short <- rowSums(counts[,1:51]) #adds up number of fragments between 100-150bp for each row/bin
  long <- rowSums(counts[,52:121])
  bins$short <- short
  bins$long <- long
  bins$ratio <- short/long
  bins$nfrags <- short+long
  
  saveRDS(data.frame(bins), paste0(output, "/counts/", sample_id, ".counts.RDS"))
}
```


```{r run fragment counts, eval = FALSE}
# filters.hg19 obtained from Delfi script 00-filtered_regions.r
filters <- load(filter_path)
samples

for(i in 1:length(samples)){
  print(samples[i])
  gr <- readRDS(paste0(output, "/granges/", samples[i], ".gr.RDS"))
  fragment_counts(sample_id = samples[i], gr = gr, output = output, filters = filters, plot_density = TRUE)
}


```



# Plot frag profiles

```{r plot individual frag profiles, eval = FALSE}
for(i in samples){
  bin.df <- readRDS(paste0(output, "/counts/", i, ".counts.RDS"))
  fp <- ggplot(bin.df, aes(start, ratio)) +
        geom_line(linewidth = 0.2) +
        labs(y = "Ratio") +
        facet_grid(.~seqnames,
                   scales = "free",
                   space = "free",
                   switch = "both") +
        scale_x_continuous(breaks = round(seq(min(bin.df$start),
                                              max(bin.df$start),
                                              by = 100000000),1)) +
        theme_minimal() +
        theme(axis.text.x = element_blank(),
              axis.ticks.x = element_blank(),
              axis.title.x = element_blank(),
              axis.line = element_line(color = "black"),
              strip.background = element_rect(
                color = "white", fill = "white"),
              strip.placement = "outside",
              strip.text.x.bottom = element_text(angle = 90),
              panel.grid.major = element_blank(),
              panel.grid.minor = element_blank())
  
ggsave(plot = fp,
       filename = paste0(output, "/graphs/", i, ".fp.png"),
       device = "png",
       width = 12, height = 1.5, units = "in")
  
}




```


Writing function to plot fragment profiles separated by group. Function requires:
* sample_ids:
  * A character vector of sample ids with fps you want to combine.
  * Samples must have corresponding count data in "output/counts" named [id].counts.RDS
* mean.colour:
  * string indicating the colour you want the mean line to be
* output:
  * path to output directory.
  * directory must contain subdirectory named counts with corresponding fp counts data (generated above)
  * directory must contain subdirectory with subdirectory called graphs
* group.name
  * a string indicating the category that the fps you are combining fall into
  * eg. if you are combining healthy control fps, enter "controls" and final plot will be called "controls.scaled.fp.png"
* scale:
  * this is automatically set to false. This option indicates whether you want the graphs to be zero-centered, or have their true median.


```{r function combine frag profiles}

combine_fps <- function(sample.ids, mean.colour, output, group.name, scale = FALSE){
  
  # combine data into one tibble
  paths <- paste0(output, "/counts/", sample.ids, ".counts.RDS")
  bins.list <- lapply(paths, readRDS)
  tib.list <- lapply(bins.list, as_tibble)
  names(tib.list) <- sample.ids
  
  if(scale == TRUE){
    for(i in 1:length(tib.list)){
      tib.list[[i]] <- mutate(tib.list[[i]], ratio = scale(ratio)[,1])
    }
  }
  combined.counts <- map2(tib.list, names(tib.list), ~ mutate(.x, id = .y)) %>%
    bind_rows() %>% dplyr::select(id, everything())
  
  if(scale == FALSE){
    saveRDS(combined.counts, paste0(output, "/counts/", group.name, ".df.RDS"))
  } else {
    saveRDS(combined.counts, paste0(output, "/counts/", group.name, ".df.scaled.RDS"))
  }
  
  # get average
  bin.averages <- combined.counts %>%
    group_by(start, seqnames) %>%
    summarise(mean.ratio = mean(ratio))
  
  # fragment profile ggplot theme
  theme_fp <- theme_minimal() +
              theme(axis.text.x = element_blank(),
                    axis.ticks.x = element_blank(),
                    axis.title.x = element_blank(),
                    axis.line = element_line(color = "black"),
                    strip.background = element_rect(color = "white", fill = "white"),
                    strip.placement = "outside",
                    strip.text.x.bottom = element_text(angle = 90),
                    panel.grid.major = element_blank(),
                    panel.grid.minor = element_blank())
  
  combined.fp.plot <- ggplot(combined.counts, aes(start, ratio)) +
      geom_line(aes(group = id),color = "#8a8884", size = 0.2) +
      labs(y = "Ratio") + 
      facet_grid(.~seqnames,
               scales = "free",
               space = "free",
               switch = "both") +
      scale_x_continuous(breaks = round(seq(min(combined.counts$start),
                                            max(combined.counts$start),
                                            by = 100000000),1)) + 
      theme_fp
  
  
  if(scale == FALSE){
    combined.fp.plot + geom_line(data = bin.averages,
                             aes(start, mean.ratio),
                             size = 0.5, color = mean.colour) + ylim(0,0.6)
    ggsave(filename = paste0(output, "/graphs/", group.name, ".unscaled.fp.png"),
           device = "png",
           width = 12,
           height = 1.5,
           units = "in")
  } else {
    combined.fp.plot + geom_line(data = bin.averages,
                             aes(start, mean.ratio),
                             size = 0.5, color = mean.colour) + ylim(-5,15)
    ggsave(filename = paste0(output, "/graphs/", group.name, ".scaled.fp.png"),
           device = "png",
           width = 12,
           height = 1.5,
           units = "in")
  }

}



```



```{r combine frag profiles unscaled}
sample_info$group
## DELFI GC
delfi_gc_info <- sample_info[sample_info$group == "delfi_gc",]

delfi_gc_samples <- delfi_gc_info$sample_id
combine_fps(sample.ids = delfi_gc_samples,
            mean.colour = "#7a1bb5",
            output = output,
            group.name = "delfi_gc")
combine_fps(sample.ids = delfi_gc_samples,
            mean.colour = "#7a1bb5",
            output = output,
            group.name = "delfi_gc",
            scale = TRUE)

## High GC - Pre-surgery
high_gc_info <- sample_info[sample_info$group == "high_gc",]

high_gc_samples <- high_gc_info$sample_id
combine_fps(sample.ids = high_gc_samples,
            mean.colour = "red",
            output = output,
            group.name = "high_gc")
combine_fps(sample.ids = high_gc_samples,
            mean.colour = "red",
            output = output,
            group.name = "high_gc",
            scale = TRUE)

# Low GC - Post-surgery
low_gc_info <- sample_info[sample_info$group == "low_gc",]

low_gc_samples <- low_gc_info$sample_id
combine_fps(sample.ids = low_gc_samples,
            mean.colour = "blue",
            output = output,
            group.name = "low_gc")
combine_fps(sample.ids = low_gc_samples,
            mean.colour = "blue",
            output = output,
            group.name = "low_gc",
            scale = TRUE)

# All GC - Pre- & Post-surgery
all_gc_info <- rbind(high_gc_info,low_gc_info)

all_gc_samples <- all_gc_info$sample_id
combine_fps(sample.ids = all_gc_samples,
            mean.colour = "#0b7d35",
            output = output,
            group.name = "all_gc")
combine_fps(sample.ids = all_gc_samples,
            mean.colour = "#0b7d35",
            output = output,
            group.name = "all_gc",
            scale = TRUE)

# DELFI Healthy

delfi_healthy_info <- sample_info[sample_info$group == "healthy",]

delfi_healthy_samples <- delfi_healthy_info$sample_id

combine_fps(sample.ids = delfi_healthy_info$sample_id,
            mean.colour = "black",
            output = output,
            group.name = "healthy")
combine_fps(sample.ids = delfi_healthy_info$sample_id,
            mean.colour = "black",
            output = output,
            group.name = "healthy",
            scale = TRUE)


```



