---
title: "03_Promoter_patterns_prop_and_quant_analysis"
author: "Judy Ann Cocadiz"
date: "2023-07-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


# Analyse "open" genes

## high prop

```{r}
all_open_genes <- vector("list", length(all_samples))
count <- 1
for(i in all_samples){
  sample_open_genes <- read_lines(paste0(output,
                                       "/promoter_patterns/potential_open_genes/all_promoters/byprop/",
                                       i, "byprop.txt"))
  all_open_genes[[count]] <- sample_open_genes
  names(all_open_genes)[count] <- i
  count <- count +1
}


```


```{r}

high_gc_open_genes <- all_open_genes[which(sample_info$group == "high_gc")]
low_gc_open_genes <- all_open_genes[which(sample_info$group == "low_gc")]
delfi_gc_open_genes <- all_open_genes[which(sample_info$group == "delfi_gc")]
healthy_open_genes <- all_open_genes[which(sample_info$group == "healthy")]


sorted_high_gc <- unlist(high_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_low_gc <- unlist(low_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_delfi_gc <- unlist(delfi_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_healthy <- unlist(healthy_open_genes) %>% table() %>%
  sort(decreasing = TRUE)

n_high_gc_samples <- which(sample_info$group == "high_gc") %>% length()
n_low_gc_samples <- which(sample_info$group == "low_gc") %>% length()
n_delfi_gc_samples  <- which(sample_info$group == "delfi_gc") %>% length()
n_healthy_samples <-  which(sample_info$group == "healthy") %>% length()



```

```{r}
sorted_all_open_genes <- unlist(all_open_genes) %>% table() %>% sort(decreasing = TRUE)
prmtrs_to_test <- names(sorted_all_open_genes[sorted_all_open_genes > 3])

```



```{r}
prmtr_yes_no <- vector("list", length(prmtrs_to_test))
count <- 1
for(i in prmtrs_to_test){
  high_gc_count <- sorted_high_gc[which(names(sorted_high_gc) == i)]
  high_gc_count <- ifelse(length(high_gc_count) > 0, high_gc_count, 0)
  low_gc_count <- sorted_low_gc[which(names(sorted_low_gc) == i)]
  low_gc_count <- ifelse(length(low_gc_count) > 0, low_gc_count, 0)
  delfi_gc_count <- sorted_delfi_gc[which(names(sorted_delfi_gc) == i)]
  delfi_gc_count <- ifelse(length(delfi_gc_count) > 0, delfi_gc_count, 0)
  healthy_count <- sorted_healthy[which(names(sorted_healthy) == i)]
  healthy_count <- ifelse(length(healthy_count) > 0, healthy_count, 0)
  
  yes.list <- c(high_gc_count,
                low_gc_count,
                delfi_gc_count,
                healthy_count)
 
  no.list <- c(n_high_gc_samples - high_gc_count,
               n_low_gc_samples - low_gc_count,
               n_delfi_gc_samples - delfi_gc_count,
               n_healthy_samples - healthy_count)
  
  prmtr_yes_no[[count]] <- tibble(yes = yes.list, no = no.list) %>% as.matrix()
  rownames(prmtr_yes_no[[count]]) <- c("high_gc", "low_gc", "delfi_gc", "healthy")
  names(prmtr_yes_no)[count] <- i
  
  count <- count + 1
}

```

```{r}

results <- tibble(gene = NA, p.value = NA)
for(i in 1:length(prmtr_yes_no)){
  gene <- names(prmtr_yes_no)[i]
  data <- prmtr_yes_no[[i]]
  test <- chisq.test(data)
  gene.results <- tibble(gene = gene, p.value = test$p.value)
  results <- bind_rows(results, gene.results)
}
results <- results[-1,]
results
p.adj <- p.adjust(results$p.value, method = p.adjust.methods, n =length(results$p.value))
results <- mutate(results, p.adj = p.adj)
results

sig_results <- results[results$p.value < 0.05,]$gene
sig_results
write_lines(sig_results, paste0(output, "/promoter_patterns/high_gc_sig_open_genes/all_promoters/high_gc_sig_open_genes_byprop.txt"))
prmtr_yes_no[names(prmtr_yes_no) %in% sig_results]

healthy_enriched <- sig_results[1:length(sig_results)]
stad_enriched <- sig_results[c(4,11,12,16,21,22,25,27,30,32,35,41)]


```
Write out enriched genes and get counts. Check if counts differ between BRCA and COAD in BRCA enriched genes.


Take genes and convert to a tsv file that can be used to create a phenogram using http://visualization.ritchielab.org/phenograms/plot
Remember to set to hg19.
To make colours the same on the graph/remove legend change "phenotype" column so that all values are the same.


```{r}
genes_to_phenogram_bm <- function(gene_list){
  mart <- useMart("ensembl")
  ds <- useDataset(dataset = "hsapiens_gene_ensembl", mart = mart)
  getBM(attributes = c("chromosome_name", "start_position",
                     "end_position", "ensembl_gene_id", "hgnc_symbol"),
        filters ="hgnc_symbol",
        values = gene_list,
        mart = ds) -> bm
  print(bm)
  bm2 <- bm %>% filter(substr(chromosome_name, 1,3) != "CHR") %>%
  mutate(seqnames = paste0("chr", chromosome_name),
         start = start_position,
         end = end_position) %>%
  dplyr::select(-c("chromosome_name", "start_position",
                   "end_position", "ensembl_gene_id")) %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
  
  # convert to hg19
  ch.path <- system.file(package = "liftOver", "extdata", "hg38ToHg19.over.chain")
  ch <- import.chain(ch.path)
  bm2 <- unlist(liftOver(bm2, ch))
  bm2 <- bm2[!duplicated(bm2$hgnc_symbol)]
  
  phen_table <- as_tibble(bm2) %>%
  mutate(seqnames = as.character(seqnames),
         chr = str_sub(seqnames, 4, nchar(seqnames))) %>%
  dplyr::select(c(chr, start, hgnc_symbol))
  colnames(phen_table) <- c("chrom", "pos", "phenotype")
  return(phen_table)
}
```


```{r}

stad_enriched_genes_by_prop <- genes_to_phenogram_bm(stad_enriched) # sometimes can't connect, just wait a few hours and try again.

write_tsv(stad_enriched_genes_by_prop,
         paste0(output, "/promoter_patterns/ready_to_phenogram/all_promoters/stad_enriched_by_prop.tsv"))
length(healthy_enriched)

healthy_enriched_genes_by_ratio <- genes_to_phenogram_bm(healthy_enriched)

nrow(healthy_enriched_genes_by_ratio)
write_tsv(healthy_enriched_genes_by_ratio,
          paste0(output, "/promoter_patterns/ready_to_phenogram/all_promoters/healthy_enriched_by_prop.tsv"))


cat(healthy_enriched)
```


# Analyse "open" genes

## high quant

```{r}
all_open_genes <- vector("list", length(all_samples))
count <- 1
for(i in all_samples){
  sample_open_genes <- read_lines(paste0(output,
                                       "/promoter_patterns/potential_open_genes/all_promoters/byquant/", i, ".byquant.txt"))
  all_open_genes[[count]] <- sample_open_genes
  names(all_open_genes)[count] <- i
  count <- count +1
}


```


```{r}

high_gc_open_genes <- all_open_genes[which(sample_info$group == "high_gc")]
low_gc_open_genes <- all_open_genes[which(sample_info$group == "low_gc")]
delfi_gc_open_genes <- all_open_genes[which(sample_info$group == "delfi_gc")]
healthy_open_genes <- all_open_genes[which(sample_info$group == "healthy")]


sorted_high_gc <- unlist(high_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_low_gc <- unlist(low_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_delfi_gc <- unlist(delfi_gc_open_genes) %>% table() %>%
  sort(decreasing = TRUE)
sorted_healthy <- unlist(healthy_open_genes) %>% table() %>%
  sort(decreasing = TRUE)

n_high_gc_samples <- which(sample_info$group == "high_gc") %>% length()
n_low_gc_samples <- which(sample_info$group == "low_gc") %>% length()
n_delfi_gc_samples  <- which(sample_info$group == "delfi_gc") %>% length()
n_healthy_samples <-  which(sample_info$group == "healthy") %>% length()



```

```{r}
sorted_all_open_genes <- unlist(all_open_genes) %>% table() %>% sort(decreasing = TRUE)
prmtrs_to_test <- names(sorted_all_open_genes[sorted_all_open_genes > 3])

```



```{r}
prmtr_yes_no <- vector("list", length(prmtrs_to_test))
count <- 1
for(i in prmtrs_to_test){
  high_gc_count <- sorted_high_gc[which(names(sorted_high_gc) == i)]
  high_gc_count <- ifelse(length(high_gc_count) > 0, high_gc_count, 0)
  low_gc_count <- sorted_low_gc[which(names(sorted_low_gc) == i)]
  low_gc_count <- ifelse(length(low_gc_count) > 0, low_gc_count, 0)
  delfi_gc_count <- sorted_delfi_gc[which(names(sorted_delfi_gc) == i)]
  delfi_gc_count <- ifelse(length(delfi_gc_count) > 0, delfi_gc_count, 0)
  healthy_count <- sorted_healthy[which(names(sorted_healthy) == i)]
  healthy_count <- ifelse(length(healthy_count) > 0, healthy_count, 0)
  
  yes.list <- c(high_gc_count,
                low_gc_count,
                delfi_gc_count,
                healthy_count)
 
  no.list <- c(n_high_gc_samples - high_gc_count,
               n_low_gc_samples - low_gc_count,
               n_delfi_gc_samples - delfi_gc_count,
               n_healthy_samples - healthy_count)
  
  prmtr_yes_no[[count]] <- tibble(yes = yes.list, no = no.list) %>% as.matrix()
  rownames(prmtr_yes_no[[count]]) <- c("high_gc", "low_gc", "delfi_gc", "healthy")
  names(prmtr_yes_no)[count] <- i
  
  count <- count + 1
}

```

```{r}

results <- tibble(gene = NA, p.value = NA)
for(i in 1:length(prmtr_yes_no)){
  gene <- names(prmtr_yes_no)[i]
  data <- prmtr_yes_no[[i]]
  test <- chisq.test(data)
  gene.results <- tibble(gene = gene, p.value = test$p.value)
  results <- bind_rows(results, gene.results)
}
results <- results[-1,]
results
p.adj <- p.adjust(results$p.value, method = p.adjust.methods, n =length(results$p.value))
results <- mutate(results, p.adj = p.adj)
results

sig_results <- results[results$p.adj < 0.05,]$gene
sig_results
write_lines(sig_results, paste0(output, "/promoter_patterns/high_gc_sig_open_genes/all_promoters/high_gc_sig_open_genes_byquant.txt"))
prmtr_yes_no[names(prmtr_yes_no) %in% sig_results]

healthy_enriched <- sig_results[1:7]
stad_enriched <- sig_results[8:length(sig_results)]


```
Write out enriched genes and get counts. Check if counts differ between BRCA and COAD in BRCA enriched genes.


Take genes and convert to a tsv file that can be used to create a phenogram using http://visualization.ritchielab.org/phenograms/plot
Remember to set to hg19.
To make colours the same on the graph/remove legend change "phenotype" column so that all values are the same.


```{r}
genes_to_phenogram_bm <- function(gene_list){
  mart <- useMart("ensembl")
  ds <- useDataset(dataset = "hsapiens_gene_ensembl", mart = mart)
  getBM(attributes = c("chromosome_name", "start_position",
                     "end_position", "ensembl_gene_id", "hgnc_symbol"),
        filters ="hgnc_symbol",
        values = gene_list,
        mart = ds) -> bm
  print(bm)
  bm2 <- bm %>% filter(substr(chromosome_name, 1,3) != "CHR") %>%
  mutate(seqnames = paste0("chr", chromosome_name),
         start = start_position,
         end = end_position) %>%
  dplyr::select(-c("chromosome_name", "start_position",
                   "end_position", "ensembl_gene_id")) %>%
  makeGRangesFromDataFrame(keep.extra.columns = TRUE)
  
  # convert to hg19
  ch.path <- system.file(package = "liftOver", "extdata", "hg38ToHg19.over.chain")
  ch <- import.chain(ch.path)
  bm2 <- unlist(liftOver(bm2, ch))
  bm2 <- bm2[!duplicated(bm2$hgnc_symbol)]
  
  phen_table <- as_tibble(bm2) %>%
  mutate(seqnames = as.character(seqnames),
         chr = str_sub(seqnames, 4, nchar(seqnames))) %>%
  dplyr::select(c(chr, start, hgnc_symbol))
  colnames(phen_table) <- c("chrom", "pos", "phenotype")
  return(phen_table)
}
```


```{r}

stad_enriched_genes_by_prop <- genes_to_phenogram_bm(stad_enriched) # sometimes can't connect, just wait a few hours and try again.

write_tsv(stad_enriched_genes_by_prop,
          paste0(output, "/promoter_patterns/ready_to_phenogram/all_promoters/stad_enriched_by_quant.tsv"))
length(healthy_enriched)

healthy_enriched_genes_by_ratio <- genes_to_phenogram_bm(healthy_enriched)

nrow(healthy_enriched_genes_by_ratio)
write_tsv(healthy_enriched_genes_by_ratio,
          paste0(output, "/promoter_patterns/ready_to_phenogram/all_promoters/healthy_enriched_by_quant.tsv"))


cat(healthy_enriched)
```

